SFJ: An implementation of Semantic Featherweight Java

### 2020-04-23 conference/camera ready/paper.tex
new place for subtyping
add typing for constants, fix typos

\makeatletter
\def\verbatim@font{\linespread{0.75}\normalfont\ttfamily}
\makeatother

% Non standard
%\usepackage{varacca}
\input semmacros

@@

    We present SFJ---Semantic Featherweight Java, an implementation of FJ which features boolean types and semantic subtyping and integrates nominal as well as structural subtyping.
    The benefits of SFJ, illustrated in the paper and the accompanying video (with audio/subtitles) \cite{UD20}, show how static typechecking of boolean types and semantic subtyping gives higher guarantees of program correctness, more flexibility and compactness of program writing.
    \keywords{Nominal subtyping \and Structural subtyping  \and Semantic Featherweight Java \and Object-oriented languages \and Boolean types \and Type theory.}
\end{abstract}

@@

Dardha \emph{et al.} \cite{Dardha2013,Dardha2017} define boolean types---based on set-theoretic operations such as \textbf{and}, \textbf{not}, \textbf{or}---and semantic subtyping for \emph{Featherweight Java} (FJ) \cite{featherweight}.
This approach allows for the integration of both nominal and structural subtyping in FJ, bringing in higher guarantees of program correctness, flexibility and compactness in program writing.
Unfortunately, these benefits were only presented at a theoretical level and not exploited practically, due to the lack of an implementation of the language and its types and type system.

@@

    Consider the set of polygons, such as triangles, squares and rhombuses given by a class hierarchy.
    We want to define a method \emph{diagonal} that takes a polygon and returns the length of its longest diagonal. This method makes sense only if the polygon passed to it has at least four sides, hence triangles are excluded.
    In Java this could be implemented in the following ways:
    $$
        \begin{array}{l}
            \synclass Polygon \ \{ \ldots\}
            \\
            \syndecl{Triangle}{Polygon} {\ldots}
            \\
            \syndecl{Other\_Polygons}{Polygon} {
                \\
                \qquad \mathbf{double}\ diagonal(Other\_Polygons \ shape)\ \{\ldots \}
                \\
                \hspace{3.5cm} \ldots
                \\
            }
            \\
            \syndecl{Square}{Other\_Polygons}{\ldots}
            \\
            \syndecl{Rhombus}{Other\_Polygons} {\ldots}
        \end{array}
    $$
    Or by means of an interface \emph{Diagonal}:
    $$
        \begin{array}{l}
            public \ \syninterface Diagonal\ \{
            \\
            \qquad \mathbf{double}\ diagonal(Polygon \ shape);
            \\
            \}
            \\
            \synclass Polygon \ \{\ldots\}
            \\
            \syndecl{Triangle}{Polygon} {\ldots}
            \\
            \synidecl{Square}{Polygon} \synimpl Diagonal \ \{\ldots\}
            \\
            \synidecl{Rhombus}{Polygon} \synimpl Diagonal \ \{\ldots\}
            \\
            \hspace{4.5cm} \ldots
        \end{array}
    $$

@@

    We propose a more elegant solution, by combining boolean types and semantic subtyping, where only static typechecking is required and we implement this in SFJ \cite{UD20}: it is enough to define a method \emph{diagonal} that has an argument of type \emph{Polygon} \textbf{and not} \emph{Triangle}, thus allowing the typechecker to check at compile time the restrictions on types:
    $$
        \begin{array}{l}
            \synclass Polygon \ \{\ldots\}
            \\
            \syndecl{Triangle}{Polygon} {\ldots}
            \\
            \syndecl{Square}{Polygon} {\ldots}
            \\
            \syndecl{Rhombus}{Polygon} {\ldots}
            \\
            \hspace{2.5cm} \ldots
            \\
            \synclass Diagonal \ \{
            \\
            %\qquad Diagonal ( )\   \{\ super();\ \}
            \hspace{2.5cm} \ldots \\
            \qquad \mathbf{double}\ diagonal((\textit{Polygon} \textbf{ and not }  \textit{Triangle}) \ shape)\{\ldots\}
            \\
            \}
        \end{array}
    $$
    We can now call \emph{diagonal} on an argument of type \emph{Polygon}: if the polygon is \textbf{not} a \emph{Triangle}, then the method returns the length of its longest diagonal; otherwise, if the polygon is a triangle, then there will be a type error at compile time.
\end{example}

@@

\section{SFJ: Design and Implementation}
\label{sec:design}
\subsection{Valid type definitions}
Since we want to use types $\tau$ in practice in SFJ, we restrict them to \emph{finite trees} whose leaves are basic types $\btypes$ \autoref{sec:terms} with no cycles.
For example, a recursive type $\alpha = [a : \alpha]$ denotes an infinite program tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)), hence we rule it out as it is not inhabitable. Similarly, the types $\alpha = [b: \beta]$, $\beta = [a = \alpha]$ create a cycle and thus would not be inhabitable.
Notice that the above types can be defined in Java. These types can be inhabited by assigning \emph{null} to all fields, but are not useful in practice.

@@

\subsection{Building semantic subtyping for SFJ}
If \autoref{alg:types} returns $True$, meaning all type definitions in an SFJ program are valid, we can proceed with building the semantic subtyping.
Leveraging the interpretation of types as sets of values to define the semantic subtyping for FJ \cite{Dardha2013,Dardha2017}, in our implementation of SFJ we keep track of the semantic subtyping relation by defining a map from a type to the set of its subtypes, satisfying the property that the set of values of a subtype is included in the set of values of the type.
We start with basic types and define $Universe$ to be the set of all types.
The mapping for basic types is defined as follows.
\begin{equation}
    \label{eq:initialSubtype}
    \begin{array}{llllll}
        & Double   & = \left\{Double, Float, Int, Short, Byte\right\} &  & Float & = \left\{Float, Short, Byte\right\} \\
        & Long     & =    \left\{Long, Int, Short, Byte\right\}       &  & Int   & = \left\{Int, Short, Byte\right\}   \\
        & Short    & =    \left\{Short, Byte\right\}                  &  & Byte  & = \left\{Byte\right\}               \\
        & Boolean  & =    \left\{Boolean\right\}                      &  & Void  & = \left\{Void\right\}               \\
        & Universe & = \{Double, Float, Long, Int,                                                                     \\
        &          & \quad Short, Byte, Boolean, Void\}
   \end{array}
\end{equation}

@@

Some comments on \autoref{alg:subtyping} will follow.
In function \textit{generateRelation} we iterate over the set of classes in an SFJ program.
If the class currently being processed contains fields or methods not present in the subtyping relation (line 5), then we add the current class to the list of \emph{unprocessed} classes (line 6), so we can process its fields and methods first.
The set of unprocessed classes will then be inspected again in a recursive call (line 10).
The next two functions of the algorithm, \emph{addClass} and \emph{checkSuperSet}, check subtyping for the current $class$ being processed and update $relation$, which is a mapping from a type to its subtypes and originally only consists of entries from \autoref{eq:initialSubtype}.
In function $addClass(class)$ we check if the type $class$ is a subtype of an existing type in $relation$ (lines 15-18), as well as the opposite, meaning if $class$ is a supertype of an existing type in $relation$ (line 19). In order to do so $checkSuperSet$ checks all fields (lines 28-39) and all methods (lines 40-51) in $class$ and compares them with an \emph{existingClass} in $relation$. If a subtyping relation is established, then it is added to $relation$ (line 53).
Finally, upon returning from $checkSuperSet$, we also add $class$ to its own relation (line 21) to satisfy reflexivity and to $Universe$ (line 22), which is a supertype of all types.

It is worth noticing that the subtyping algorithm finds all nominal and structural subtypes of a given type. This is due to the fact that all pairs of types are inspected.
Recall from \autoref{sec:intro} that nominal subtyping is name-based and given by the class hierarchy defined by the programmer, whether structural subtyping is structure-based and given by the set-inclusion of fields and methods. In particular, it is contra-variant with respect to this set-inclusion.
\autoref{alg:subtyping} finds all structural subtypes of a given class because it checks that its fields and methods are a superset of existing types in $relation$.
On the other hand, it also finds all nominal subtypes because a class inherits all fields and methods of its superclass and as such its fields and methods are a superset of its superclass.
This means that checking for structural subtyping is enough because nominal subtyping will be captured due to inheritance of fields and methods.

Finally, a note on complexity.
The complexity of \autoref{alg:types} is $\mathcal{O}(n)$, and the complexity of \autoref{alg:subtyping} is $\mathcal{O}(n^{2})$, with $n$ being the size of the input.
The reason for an exponential complexity \autoref{alg:subtyping} is due to the symmetric check of structural subtyping between a $class$ and an $existingClass$ in $relation$.
Notice that if we were to only work with nominal subtyping, then we only require traversing the class hierarchy once,  which gives an $\mathcal{O}(n)$ complexity.

\subsection{Type system for SFJ}
The subtyping relation generated by \autoref{alg:subtyping} is used in the type system for SFJ terms, given in \autoref{sec:terms}.
The complete type system and its properties are presented in relevant work \cite{Dardha2013,Dardha2017} and are beyond the scope of this paper.
Let us briefly comment on typing expressions in SFJ.
Field access $e.a$, method call $e.m(e)$ and object creation $\synnew C(\wt{e})$ are typed in the same way as in Java.
Let us comment on constants...

%
\begin{algorithm}
    \caption{Semantic Subtyping for SFJ Classes}
    \label{alg:subtyping}

@@

            \If{$flag = True$}{
                $relation[other].add$
            }
        }
    }
\end{algorithm}
%
\subsection{Polygons: Continued}
Let us illustrate the subtyping algorithm with our \emph{Polygons} \autoref{polygons}.
The algorithm generates the following subtyping relation, which also includes the subtyping relation for basic types defined in the first step.
\begin{flalign*}
    &Polygon = \left\{Polygon, Triangle, Square, Rhombus \right\} \qquad Triangle = \left\{Triangle\right\}\\
    &Square = \left\{Square\right\} \qquad  Rhombus  = \left\{Rhombus\right\} \qquad
    Diagonal = \left\{Diagonal\right\}
\end{flalign*}
Recall method \emph{diagonal} in class \emph{Diagonal}, we can see that the result of the set operation on its parameter type gives the following set of polygons:
\begin{flalign*}
    Polygon \textbf{ and not } Triangle = \left\{Polygon, Square, Rhombus \right\}
\end{flalign*}
If we pass to \emph{diagonal} an argument of type \emph{Triangle}, as it is not contained in the computed set of subtypes, we get a typechecking error at compile time \cite{UD20}.

The subtyping algorithm finds all nominal and structural subtypes.
It finds all nominal subtypes as they inherit all fields and methods of their super class, so are always guaranteed to be a superset.
Similarly, given all pairs of types, the algorithm checks if one is a superset of the other, i.e.,\ they share common fields and methods, and if so, they are related by structural subtyping.
For example, type $empty = []$, will have all classes as structural subtypes.
%The above algorithm will without differentiating, find all nominal and structural subtypes as we always consider all other classes in the program to be potential subtype candidates.

@@

@@

\section{Related Work and Conclusion}
\label{sec:conclusion}
Semantic subtyping approach goes back to more than two decades ago \cite{Aiken,Damm}.
Notable lines of work include:
Hosoya and Pierce \cite{XML1,XML2,XML3} who define XDuce, an XML-oriented language designed specifically to transform XML documents in other XML documents satisfying certain properties.
Castagna \emph{et al.} \cite{gentle,Cas05,FCB08} extend XDuce with first-class functions and arrow types and implement it as $\mathbb{C}$Duce. The starting point of their framework is a higher-order $\lambda-$calculus with pairs and projections.
Muehlboeck and Tate \cite{Muehlboeck2018} define a syntactic framework with boolean connectives which has been implemented in the Ceylon programming language \cite{Ceylon2016}.
Castagna \emph{et al.} \cite{Cpi} define $\mathbb{C}\pi$, a variant of the asynchronous $\pi-$calculus \cite{sangiorgi}, where channel types are augmented with boolean connectives.
Ancona and Lagorio~\cite{AL10} define subtyping for infinite types coinductively by using union and object type constructors, where types are interpreted as sets of value of the language.
Bonsangue \emph{et al.}~\cite{BRABR14} study a coalgebraic approach to coinductive types and define a set-theoretic interpretation of coinductive types with union types.
Pearce \cite{Pearce19} defines semantic subtyping for rewriting rules in the Whiley Rewrite Language and for a flow-typing calculus  \cite{Pearce13}.

>>>>

% Non standard
%\usepackage{varacca}
\input{semmacros}

@@

    We present SFJ---Semantic Featherweight Java, an implementation of FJ which features boolean types, semantic subtyping and integrates nominal as well as structural subtyping.
    The benefits of SFJ, illustrated in the paper and the accompanying video (with audio/subtitles) \cite{UD20}, show how static typechecking of boolean types and semantic subtyping gives higher guarantees of program correctness, more flexibility and compactness of program writing.
    \keywords{Nominal subtyping \and Structural subtyping  \and Semantic Featherweight Java \and Object-oriented languages \and Boolean types \and Type theory.}
\end{abstract}

@@

Dardha \emph{et al.} \cite{Dardha2013,Dardha2017} define boolean types---based on set-theoretic operations such as \textbf{and}, \textbf{not}, \textbf{or}---and semantic subtyping for \emph{Featherweight Java} (FJ) \cite{featherweight}.
This approach allows for the integration of both nominal and structural subtyping in FJ, bringing in higher guarantees of program correctness, flexibility and compactness in program writing.
Unfortunately, these benefits were only presented at a theoretical level and not exploited practically, due to the lack of an implementation of the language, its types and type system.

@@

    Consider the set of polygons, such as triangles, squares and rhombuses given by a class hierarchy.
    We want to define a method \emph{diagonal} that takes a polygon and returns the length of its longest diagonal.
    This method makes sense only if the polygon passed to it has at least four sides, hence triangles are excluded.
    In Java this could be implemented in the following ways:

    \begin{equation*}
        \begin{array}{l}
            \synclass Polygon \ \{ \ldots\}
            \\
            \syndecl{Triangle}{Polygon} {\ldots}
            \\
            \syndecl{Other\_Polygons}{Polygon} {
                \\
                \qquad \mathbf{double}\ diagonal(Other\_Polygons \ shape)\ \{\ldots \}
                \\
                \hspace{3.5cm} \ldots
                \\
            }
            \\
            \syndecl{Square}{Other\_Polygons}{\ldots}
            \\
            \syndecl{Rhombus}{Other\_Polygons} {\ldots}
        \end{array}
    \end{equation*}

    Or by means of an interface \emph{Diagonal}:

    \begin{equation*}
        \begin{array}{l}
            public \ \syninterface Diagonal\ \{
            \\
            \qquad \mathbf{double}\ diagonal(Polygon \ shape);
            \\
            \}
            \\
            \synclass Polygon \ \{\ldots\}
            \\
            \syndecl{Triangle}{Polygon} {\ldots}
            \\
            \synidecl{Square}{Polygon} \synimpl Diagonal \ \{\ldots\}
            \\
            \synidecl{Rhombus}{Polygon} \synimpl Diagonal \ \{\ldots\}
            \\
            \hspace{4.5cm} \ldots
        \end{array}
    \end{equation*}

@@

    We propose a more elegant solution, by combining boolean types and semantic subtyping, where only static typechecking is required and we implement this in SFJ \cite{UD20}: it is enough to define a method \emph{diagonal} that has an argument of type \emph{Polygon} \textbf{and not} \emph{Triangle}, thus allowing the typechecker to check at compile time the restrictions on types:

    \begin{equation*}
        \begin{array}{l}
            \synclass Polygon \ \{\ldots\}
            \\
            \syndecl{Triangle}{Polygon} {\ldots}
            \\
            \syndecl{Square}{Polygon} {\ldots}
            \\
            \syndecl{Rhombus}{Polygon} {\ldots}
            \\
            \hspace{2.5cm} \ldots
            \\
            \synclass Diagonal \ \{
            \\
            %\qquad Diagonal ( )\   \{\ super();\ \}
            \hspace{2.5cm} \ldots \\
            \qquad \mathbf{double}\ diagonal((\textit{Polygon} \textbf{ and not }  \textit{Triangle}) \ shape)\{\ldots\}
            \\
            \}
        \end{array}
    \end{equation*}

    We can now call \emph{diagonal} on an argument of type \emph{Polygon}: if the polygon is \textbf{not} a \emph{Triangle}, then the method returns the length of its longest diagonal; otherwise, if the polygon is a triangle, then there will be a type error at compile time.
\end{example}

@@

\section{SFJ: Design and Implementation}
\label{sec:design}
\subsection{Valid type definitions}
Since we want to use types $\tau$ in practice in SFJ, we restrict them to \emph{finite trees} whose leaves are basic types $\btypes$ \autoref{sec:terms} with no cycles.
For example, a recursive type $\alpha = [a : \alpha]$ denotes an infinite program tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)), hence we rule it out as it is not inhabitable.
Similarly, the types $\alpha = [b: \beta]$, $\beta = [a = \alpha]$ create a cycle and thus would not be inhabitable.
Notice that the above types can be defined in Java as these types can be inhabited by assigning \emph{null} to all fields, but are not useful in practice.

@@

\subsection{Building semantic subtyping for SFJ}

If \autoref{alg:types} returns $True$, meaning all type definitions in an SFJ program are valid, we can proceed with building the semantic subtyping.
Leveraging the interpretation of types as sets of values to define the semantic subtyping for FJ \cite{Dardha2013,Dardha2017}, in our implementation of SFJ we keep track of the semantic subtyping relation by defining a map from a type to the set of its subtypes, satisfying the property that the set of values of a subtype is included in the set of values of the type.
We start with basic types and define $Universe$ to be the set of all types.
The mapping for basic types is defined as follows.

\begin{equation}
    \label{eq:initialSubtype}
    \begin{array}{llllll}
        & Double   & = \left\{Double, Float, Int, Short, Byte\right\} &  & Float & = \left\{Float, Short, Byte\right\} \\
        & Long     & =    \left\{Long, Int, Short, Byte\right\}       &  & Int   & = \left\{Int, Short, Byte\right\}   \\
        & Short    & =    \left\{Short, Byte\right\}                  &  & Byte  & = \left\{Byte\right\}               \\
        & Boolean  & =    \left\{Boolean\right\}                      &  & Void  & = \left\{Void\right\}               \\
        & Universe & = \{Double, Float, Long, Int,                                                                     \\
        &          & \quad Short, Byte, Boolean, Void\}
   \end{array}
\end{equation}

@@

Some comments on \autoref{alg:subtyping} will follow.
In function $generateRelation$ we iterate over the set of classes in an SFJ program.
If the class currently being processed contains types in its fields or methods not present in the subtyping relation (line 5, 30, 42), then we add the current class to the list of \emph{unprocessed} classes (line 6) so we can process its later as it is currently undecidable.
The set of unprocessed classes will then be inspected again in a recursive call (line 10).
The next two functions of the algorithm, \emph{addClass} and \emph{checkSuperSet}, check subtyping for the current $class$ being processed and update $relation$, which is a mapping from a type to its subtypes and originally only consists of entries from \autoref{eq:initialSubtype}.
In function $addClass(class)$ we check if the type $class$ is a subtype of an existing type in $relation$ (lines 15-18), as well as the opposite, meaning if $class$ is a supertype of an existing type in $relation$ (line 19).
In order to do so $checkSuperSet$ checks all fields (lines 28-39) and all methods (lines 40-51) in $class$ and compares them with an \emph{existingClass} in $relation$.
If a subtyping relation is established, then it is added to $relation$ (line 53).
Finally, upon returning from $checkSuperSet$, we also add $class$ to its own relation (line 21) to satisfy reflexivity and to $Universe$ (line 22), which is a supertype of all types.

\begin{algorithm}
    \caption{Semantic Subtyping for SFJ Classes}
    \label{alg:subtyping}

@@

            \If{$flag = True$}{
                $relation[other].add$
            }
        }
    }
\end{algorithm}

It is worth noticing that the subtyping algorithm finds all nominal and structural subtypes of a given type.
This is due to the fact that all pairs of types are inspected.
Recall from \autoref{sec:intro} that nominal subtyping is name-based and given by the class hierarchy defined by the programmer, whereas structural subtyping is structure-based and given by the set-inclusion of fields and methods.
In particular, it is contra-variant with respect to this set-inclusion.
\autoref{alg:subtyping} finds all structural subtypes of a given class because it checks that its fields and methods are a superset of existing types in $relation$.
On the other hand, it also finds all nominal subtypes because a class inherits all fields and methods of its superclass and as such its fields and methods are a superset of its superclass.
This means that checking for structural subtyping is enough because nominal subtyping will be captured due to inheritance of fields and methods.

Finally, a note on complexity.
The complexity of \autoref{alg:types} is $\mathcal{O}(n)$, and the complexity of \autoref{alg:subtyping} is $\mathcal{O}(n^{2})$, with $n$ being the size of the input.
The reason for a quadratic complexity \autoref{alg:subtyping} is due to the symmetric check of structural subtyping between a $class$ and an $existingClass$ in $relation$.
Notice that if we were to only work with nominal subtyping, then we only require traversing the class hierarchy once,  which gives an $\mathcal{O}(n)$ complexity.

\subsection{Type system for SFJ}

The subtyping relation generated by \autoref{alg:subtyping} is used in the type system for SFJ terms, given in \autoref{sec:terms}.
The complete type system and its properties are presented in relevant work \cite{Dardha2013,Dardha2017} and are beyond the scope of this paper.
Let us briefly comment on typing expressions in SFJ.
Field access $e.a$, method call $e.m(e)$, variables $x$ and object creation $\synnew C(\wt{e})$ are typed in the same way as in Java.

For the constants $c$, these are the literal values in our language such as $42$, $3.14$ or $True$.
In most languages such as Java, literal values are cast to the type that is expected in the context they are used.
For example, a method $\textbf{int}\ getNum()\{return\ 42\}$ automatically casts the value to the type $int$.

However, in our system, types represent sets of values which must interpret boolean connectives in a set-theoretic way.
Given the boolean type $int\ \textbf{and not}\ byte$, this type represents the values $-2^31$ to $2^31 - 1$ from which are removed the values $-128$ to $127$.
If we were to type the value $42$ as an integer, this would be accepted by the boolean type, but would not be in the set of values of the type, thus breaking our semantic type system.
Therefore, for all constants $c$, we use a strong typing to assign values the most restrictive types possible, i.e. the type with the smallest amount of values which still contains this value.
With this typing rule, the value $42$ is assigned the type $Byte$, so would correctly not be accepted as a value of the type $(int\ \textbf{and not}\ byte)$.

We have now defined how to type any expression that can be formed by the syntax and so have defined our typing rules.

\subsection{Polygons: Continued}

Let us illustrate the subtyping algorithm with our \emph{Polygons} \autoref{polygons}.
The algorithm generates the following subtyping relation, which also includes the subtyping relation for basic types defined in the first step.

\begin{flalign*}
    &Polygon = \left\{Polygon, Triangle, Square, Rhombus \right\} \qquad Triangle = \left\{Triangle\right\}\\
    &Square = \left\{Square\right\} \qquad  Rhombus  = \left\{Rhombus\right\} \qquad
    Diagonal = \left\{Diagonal\right\}
\end{flalign*}

Recall method \emph{diagonal} in class \emph{Diagonal}, we can see that the result of the set operation on its parameter type gives the following set of polygons:

\begin{flalign*}
    Polygon \textbf{ and not } Triangle = \left\{Polygon, Square, Rhombus \right\}
\end{flalign*}

If we pass to \emph{diagonal} an argument of type \emph{Triangle}, as it is not contained in the computed set of subtypes, we get a typechecking error at compile time \cite{UD20}.

%It finds all nominal subtypes as they inherit all fields and methods of their super class, so are always guaranteed to be a superset.
%Similarly, given all pairs of types, the algorithm checks if one is a superset of the other, i.e.,\ they share common fields and methods, and if so, they are related by structural subtyping.
%For example, type $empty = []$, will have all classes as structural subtypes.
%The above algorithm will without differentiating, find all nominal and structural subtypes as we always consider all other classes in the program to be potential subtype candidates.

\subsection{Nominal vs Structural}

@@

\subsection{Method Types}

@@

\section{Related Work and Conclusion}
\label{sec:conclusion}

Semantic subtyping approach goes back to more than two decades ago \cite{Aiken,Damm}.
Notable lines of work include:
Hosoya and Pierce \cite{XML1,XML2,XML3} who define XDuce, an XML-oriented language designed specifically to transform XML documents in other XML documents satisfying certain properties.
Castagna \emph{et al.} \cite{Castagna2005,Cas05,FCB08} extend XDuce with first-class functions and arrow types and implement it as $\mathbb{C}$Duce. The starting point of their framework is a higher-order $\lambda-$calculus with pairs and projections.
Muehlboeck and Tate \cite{Muehlboeck2018} define a syntactic framework with boolean connectives which has been implemented in the Ceylon programming language \cite{Ceylon2016}.
Castagna \emph{et al.} \cite{Cpi} define $\mathbb{C}\pi$, a variant of the asynchronous $\pi-$calculus \cite{sangiorgi}, where channel types are augmented with boolean connectives.
Ancona and Lagorio~\cite{AL10} define subtyping for infinite types coinductively by using union and object type constructors, where types are interpreted as sets of value of the language.
Bonsangue \emph{et al.}~\cite{BRABR14} study a coalgebraic approach to coinductive types and define a set-theoretic interpretation of coinductive types with union types.
Pearce \cite{Pearce19} defines semantic subtyping for rewriting rules in the Whiley Rewrite Language and for a flow-typing calculus  \cite{Pearce13}.
